DATA 621 Homework #1

### 2019-09-25 hw1/doc/HW1.Rmd

### Missing Values

During our first look at the data it was noted that there were variables that are missing data.  Here's a look at what variables are missing data and how big of a problem is it:

@@

The hit by pitcher varriable is missing over 90% of it's data.  Based on the weak correlation with the response variable we will exclude it from consideration in our model.

@@

### Zero Values

@@

This isn't nearly as large of a problem as the missing values.  Now it's time to fix the data issues.

@@

The range of data looks good except for a single zero value. Since there are no recorded seasons with zero wins in MLB, we should impute a value in it's place. Here the mean (`r round(mean(training$TARGET_WINS), 0)`) seems as good as any value, so we will replace it.

```{r}
# Replace 0 with the mean value
training[which(training$TARGET_WINS == 0), "TARGET_WINS"] <- round(mean(training$TARGET_WINS), 0)
```

@@

We can replace these values with imputed values, but the distribution looks to be bimodal, so using a mean or median (which is squarely between those peaks) may cause some issues with the model. So, instead, we will impute values using regression.

We can impute a value for this variable by looking at it's nearest neighbors (based on other variables) and taking a weighted average of their values.

```{r}
# Change 0's to NA so they too can be imputed
training <- training %>%
  mutate(TEAM_BATTING_SO = ifelse(TEAM_BATTING_SO == 0, NA, TEAM_BATTING_SO))

knn <- training %>% knnImputation()
impute_me <- is.na(training$TEAM_BATTING_SO)
training[impute_me,"TEAM_BATTING_SO"] <- knn[impute_me,"TEAM_BATTING_SO"] 
```

Looking at our distribution, the shape hasn't changed radically and the means and medians ony shifted slightly:

```{r}
training %>% 
  ggplot(aes(TEAM_BATTING_SO)) + 
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = mean(TEAM_BATTING_SO, na.rm = T)), col = "red", lty = 2) +
  geom_vline(aes(xintercept = median(TEAM_BATTING_SO, na.rm = T)), col = "green", lty = 2) +
  labs(x = element_blank(),
       y = "Count",
       title = "Distribution of Strikeouts (Batter)",
       subtitle = "After Imputation",
       caption = "* Red line is the mean value and green is the median")
```

@@

The zeroes may be legitimate here so we will leave them alone. For the NAs, we can use the same KNN imputation we used above for strikeouts

```{r}
impute_me <- is.na(training$TEAM_BASERUN_SB)
training[impute_me,"TEAM_BASERUN_SB"] <- knn[impute_me,"TEAM_BASERUN_SB"]

training %>% 
  ggplot(aes(TEAM_BASERUN_SB)) + 
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = mean(TEAM_BASERUN_SB, na.rm = T)), col = "red", lty = 2) +
  geom_vline(aes(xintercept = median(TEAM_BASERUN_SB, na.rm = T)), col = "green", lty = 2) +
  labs(x = element_blank(),
       y = "Count",
       title = "Distribution of Stolen Bases",
       subtitle = "After Imputation",
       caption = "* Red line is the mean value and green is the median")
```

Again we see little structural change in our distribution.

@@

```{r}
impute_me <- is.na(training$TEAM_BASERUN_CS)
training[impute_me,"TEAM_BASERUN_CS"] <- knn[impute_me,"TEAM_BASERUN_CS"] 
```

@@

We *could* make an assumption that these are all cases where there were no batters hit by a pitch, but that seems ill-advised given the distribution above. In this case, we might be wise to leave them as NA for now.

@@

We can use the KNN imputation to help with the NA values

```{r}
impute_me <- is.na(training$TEAM_PITCHING_SO)
training[impute_me,"TEAM_PITCHING_SO"] <- knn[impute_me,"TEAM_PITCHING_SO"] 
```

With the NA's fixed, we can turn our attention now to the extreme values.

@@

We'll make a high-yet-reasonable assumption of a mean 11 innings per game, and call anything more than 5,346 strikeouts an invalid data point by setting them to NA so they will be disregarded in any modeling.

```{r}
training[which(training$TEAM_PITCHING_SO > 5346),"TEAM_PITCHING_SO"] <- NA

training %>% 
  ggplot(aes(TEAM_PITCHING_SO)) + 
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = mean(TEAM_PITCHING_SO, na.rm = T)), col = "red", lty = 2) +
  geom_vline(aes(xintercept = median(TEAM_PITCHING_SO, na.rm = T)), col = "green", lty = 2) +
  labs(x = element_blank(),
       y = "Count",
       title = "Distribution of Strikeouts Pitched",
       subtitle = "After Removing Invalid Data and Imputation",
       caption = "* Red line is the mean value and green is the median")
```

@@

Again, we use the KNN imputation from earlier to fill in NAs with imputed values.

```{r}
impute_me <- is.na(training$TEAM_FIELDING_DP)
training[impute_me,"TEAM_FIELDING_DP"] <- knn[impute_me,"TEAM_FIELDING_DP"] 
```

@@

```{r}
set.seed(42)
train_index <- createDataPartition(training$TARGET_WINS, p = .8, list = FALSE, times = 1)

@@

@@

```{r}
fit <- lm(TARGET_WINS ~ ., moneyball_train)
summary(fit)
```

```{r}
moneyball_test$y_hat <- predict(fit, moneyball_test)
moneyball_test <- moneyball_test %>%
  mutate(error = TARGET_WINS - y_hat) %>%
  mutate(squared_error = error ^ 2)

ggplot(moneyball_test, aes(error)) +
  geom_histogram(bins = 50)

summary(moneyball_test$squared_error)
```

## Appendix

>>>>

### Strange Data Values

#### Missing Values

During our first look at the data it was noted that there were variables that are missing data.  Here's a look at what variables are missing data and how big of a problem it is:

@@

The hit by pitcher varriable is missing over 90% of it's data.  We will exclude it from consideration in our model.

@@

#### Zero Values

@@

This isn't nearly as large of a problem as the missing values.

### Deeper Dive into the Variables

@@

The range of data looks good.  There is a single zero value, which is the Washington Nationals 1872 season.

@@

We will replace these values with imputed values, but the distribution looks to be bimodal, so using a mean or median (which is squarely between those peaks) may cause some issues with the model. So, instead, we will impute values using regression.

We will impute values for this variable by looking at it's nearest neighbors (based on other variables) and taking a weighted average of their values.

@@

The zeroes may be legitimate here so we will leave them alone. For the NAs, we will use the same KNN imputation we used above for strikeouts to impute values.

@@

@@

We will drop this variable as we cannot responsibly impute values with such thin data.

@@

@@

We'll make a high-yet-reasonable assumption of a mean 11 innings per game.  We will call anything more than 5,346 strikeouts an invalid data point by setting them to NA so they will be imputed prior to modeling.

@@

Again, we will use the KNN imputation from earlier to fill in NAs with imputed values.

First we will remove the invalid data and prep it for imputation.  We will drop the hit by pitcher variable from the dataset.

```{r}
remove_bad_values <- function(df){
  # Change 0's to NA so they too can be imputed
  df <- df %>%
  mutate(TEAM_BATTING_SO = ifelse(TEAM_BATTING_SO == 0, NA, TEAM_BATTING_SO))
  # Remove the high pitching strikeout values
  df[which(df$TEAM_PITCHING_SO > 5346),"TEAM_PITCHING_SO"] <- NA
  # Drop the hit by pitcher variable
  df %>%
    select(-TEAM_BATTING_HBP)
}

training <- remove_bad_values(training)
evaluation <- remove_bad_values(evaluation)
```

```{r}
set.seed(42)
knn <- training %>% knnImputation()

impute_values <- function(df, knn){
  impute_me <- is.na(df$TEAM_BATTING_SO)
  df[impute_me,"TEAM_BATTING_SO"] <- knn[impute_me,"TEAM_BATTING_SO"] 
  impute_me <- is.na(df$TEAM_BASERUN_SB)
  df[impute_me,"TEAM_BASERUN_SB"] <- knn[impute_me,"TEAM_BASERUN_SB"] 
  impute_me <- is.na(df$TEAM_BASERUN_CS)
  df[impute_me,"TEAM_BASERUN_CS"] <- knn[impute_me,"TEAM_BASERUN_CS"] 
  impute_me <- is.na(df$TEAM_PITCHING_SO)
  df[impute_me,"TEAM_PITCHING_SO"] <- knn[impute_me,"TEAM_PITCHING_SO"]
  impute_me <- is.na(df$TEAM_FIELDING_DP)
  df[impute_me,"TEAM_FIELDING_DP"] <- knn[impute_me,"TEAM_FIELDING_DP"]
  return(df)
}

training <- impute_values(training, knn)
```

### Feature Engineering

The batting singles is not included but we can back it out of the hits.  We will do this.

```{r}
add_features <- function(df){
  df %>%
    mutate(TEAM_BATTING_1B = TEAM_BATTING_H - TEAM_BATTING_2B - TEAM_BATTING_3B - TEAM_BATTING_HR)
}

training <- add_features(training)
evaluation <- add_features(evaluation)
```

### Results

Here's what the data look like after imputation and correction:

```{r warning=FALSE}
training %>%
  gather(variable, value) %>%
  ggplot(., aes(value)) + 
  geom_density(fill = "indianred4", color="indianred4") + 
  facet_wrap(~variable, scales ="free", ncol = 4) +
  labs(x = element_blank(), y = element_blank())
```

@@

We will divide the data into training and test sets using a 70/30 split.  We will build our models on the training set and evaluate it on the test set.

```{r}
set.seed(42)
train_index <- createDataPartition(training$TARGET_WINS, p = .7, list = FALSE, times = 1)

@@

### Kitchen Sink Model

We will begin with a "kitchen sink" model.

```{r}
kitchen_sink <- lm(TARGET_WINS ~ ., moneyball_train)
summary(kitchen_sink)
```

It does a fairly good job predicting, but there are a lot of variables that are not statistically significant.

### Simple Model

Let's try to create a simplier model. We will pick variables that had high correlations and exclude the pitching variables which would introduce autocorrelation issues.

```{r}
simple_fit <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_BB + TEAM_FIELDING_E, data = moneyball_train)
summary(simple_fit)
```

This model isn't as fitted to the data as well but the variables are statistically significant.

### Higher Order Stepwise Regression

For the third model we will use a stepwise regression method using a backwards elimination process.  We also introduce some higher order polynomial variables.

```{r}
full_formula <- "TARGET_WINS ~ TEAM_BATTING_2B + TEAM_BATTING_3B + TEAM_BATTING_HR + TEAM_BATTING_BB + TEAM_BATTING_SO + TEAM_BASERUN_SB + TEAM_BASERUN_CS + TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_PITCHING_BB + TEAM_PITCHING_SO + TEAM_FIELDING_E + TEAM_FIELDING_DP + TEAM_BATTING_1B + I(TEAM_BATTING_2B^2) + I(TEAM_BATTING_3B^2) + I(TEAM_BATTING_HR^2) + I(TEAM_BATTING_BB^2) + I(TEAM_BATTING_SO^2) + I(TEAM_BASERUN_SB^2) + I(TEAM_BASERUN_CS^2) + I(TEAM_PITCHING_H^2) + I(TEAM_PITCHING_HR^2) + I(TEAM_PITCHING_BB^2) + I(TEAM_PITCHING_SO^2) + I(TEAM_FIELDING_E^2) + I(TEAM_FIELDING_DP^2) + I(TEAM_BATTING_1B^2) + I(TEAM_BATTING_2B^3) + I(TEAM_BATTING_3B^3) + I(TEAM_BATTING_HR^3) + I(TEAM_BATTING_BB^3) + I(TEAM_BATTING_SO^3) + I(TEAM_BASERUN_SB^3) + I(TEAM_BASERUN_CS^3) + I(TEAM_PITCHING_H^3) + I(TEAM_PITCHING_HR^3) + I(TEAM_PITCHING_BB^3) + I(TEAM_PITCHING_SO^3) + I(TEAM_FIELDING_E^3) + I(TEAM_FIELDING_DP^3) + I(TEAM_BATTING_1B^3) + I(TEAM_BATTING_2B^4) + I(TEAM_BATTING_3B^4) + I(TEAM_BATTING_HR^4) + I(TEAM_BATTING_BB^4) + I(TEAM_BATTING_SO^4) + I(TEAM_BASERUN_SB^4) + I(TEAM_BASERUN_CS^4) + I(TEAM_PITCHING_H^4) + I(TEAM_PITCHING_HR^4) + I(TEAM_PITCHING_BB^4) + I(TEAM_PITCHING_SO^4) + I(TEAM_FIELDING_E^4) + I(TEAM_FIELDING_DP^4) + I(TEAM_BATTING_1B^4)"

full_model <- lm(full_formula, moneyball_train)
step_back <- MASS::stepAIC(full_model, direction="backward", trace = F)
poly_call <- summary(step_back)$call
step_back <- lm(poly_call[2], moneyball_train)
summary(step_back)
```

This model has the highest adjusted R-squared but has some variables that are statistically insignificant.

@@

In order to select which model is the "best" we will test it against a test set.  We will examine the difference between the predicted and actual values.

### Kitchen Sink Results

```{r}
moneyball_test$kitchen_sink <- predict(kitchen_sink, moneyball_test)
moneyball_test <- moneyball_test %>%
  mutate(kitchen_sink_error = TARGET_WINS - kitchen_sink)

ggplot(moneyball_test, aes(kitchen_sink_error)) +
  geom_histogram(bins = 50)

summary(moneyball_test$kitchen_sink_error)
```

### Stepwise Results

```{r}
moneyball_test$step_back <- predict(step_back, moneyball_test)
moneyball_test <- moneyball_test %>%
  mutate(step_back_error = TARGET_WINS - step_back)

moneyball_test %>%
  filter(step_back_error > -100) %>%
  ggplot(., aes(step_back_error)) +
  geom_histogram(bins = 50) +
  labs(caption = "Outlier removed")

summary(moneyball_test$step_back_error)
```

This model did not preform very well on the out of sample data.

### Simple Model

```{r}
moneyball_test$simple <- predict(simple_fit, moneyball_test)
moneyball_test <- moneyball_test %>%
  mutate(simple_error = TARGET_WINS - simple)

ggplot(moneyball_test, aes(simple_error)) +
  geom_histogram(bins = 50)

summary(moneyball_test$simple_error)
```

[^1]: https://en.wikipedia.org/wiki/Base_on_balls
